Weighted Model Counting with Conditional Weights

### 2020-08-05 paper/paper.tex
Improve the introduction

% TODO
Introduce WMC \cite{DBLP:conf/aaai/SangBK05}

Over the last fifteen years, WMC has been extended and generalised in many ways,
e.g., to handle continuous probability distributions
\cite{DBLP:conf/ijcai/BellePB15}, first-order probabilistic theories
\cite{DBLP:conf/ijcai/BroeckTMDR11,DBLP:journals/cacm/GogateD16}, and infinite
domains \cite{DBLP:conf/aaai/Belle17}. Furthermore, WMC has been generalised
from weights in the interval $[0, 1]$ or positive real numbers to an arbitrary
semiring, making it applicable to domains such as parsing, dynamic programming,
constraint programming, databases, networks, etc.
\cite{DBLP:journals/japll/KimmigBR17}. It has been applied to Bayesian networks
\cite{DBLP:conf/aaai/SangBK05,DBLP:books/daglib/0024906} and probabilistic
graphical models more generally \cite{DBLP:conf/ecsqaru/ChoiKD13}, probabilistic
logic programming languages \cite{DBLP:journals/tplp/FierensBRSGTJR15} and
probabilistic programs with discrete probability distributions
\cite{DBLP:journals/corr/abs-2005-09089}. Many WMC algorithms have been proposed
based primarily either on knowledge compilation
\cite{DBLP:conf/ecai/Darwiche04,DBLP:conf/ijcai/OztokD15,DBLP:conf/ijcai/LagniezM17}
or search \cite{DBLP:conf/aaai/SangBK05}.

The most well-studied version of WMC assigns weights to models based on
weights on literals, i.e., the weight of a model is the product of the weights
of all literals in it. This severely restricts what probability distributions
can be represented, and this limitation is usually overcome by increasing the
size of the problem, i.e., adding more literals.

We show that while this workaround is always possible, it can be significantly
more efficient to consider \emph{conditional weights} on literals. Similarly to
conditional probabilities, conditional weights allow us to represent
probabilities in a manner that is both efficient and interpretable. Not having
additional variables that are only needed to manipulate the representation into
a specific format helps with both efficiency and interpretability. Similarly to
Bayesian networks, the size of the representation scales with independence and
literal-weight WMC is a special case when everything is independent.

We demonstrate a specific example of this by presenting a new way to
encode Bayesian networks into instances of WMC and adapting a WMC algorithm
ADDMC to run on the new format. ADDMC \cite{DBLP:conf/aaai/DudekPV20} is a
recently-proposed algorithm for WMC based on manipulating functions on Boolean
algebras using an efficient representation for such functions known as
\emph{algebraic decision diagrams} (ADDs)
\cite{DBLP:journals/fmsd/BaharFGHMPS97}. We show that with a few minor
modifications the algorithm can be adapted to work on our new WMC format

Showing that the new encoding produces an asymptotically smaller problem (w.r.t.
the numbers of literals and ADDs) and results in faster inferences.

% TODO: an overview of all results, with forward references

@@

\section{WMC as a Measure}

@@

Finally, a measure $\mu\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ is
\emph{factorable} if there exists a factored weight function $\nu\colon 2^U \to
\mathbb{R}_{\ge 0}$ that induces $\mu$. In this formulation, the process of
calculating the value of $\mu_\nu(x)$ for some $x \in 2^{2^U}$ with a given
definition of $\nu$ is known as WMC.

@@

\section{Encoding Bayesian Networks Using Conditional Weights}

@@

\correctness*
\begin{proof}
  Let $\mathcal{V} = \{ X, Y_1, \dots, Y_n \}$. Then
  \begin{align*}
    (\exists_U (\phi \cdot [\lambda_{X=x}]))(\emptyset) &= \sum_{T \in 2^U} (\phi \cdot [\lambda_{X=x}])(T) \\
                                                        &= \sum_{\lambda_{X=x} \in T \in 2^U} \phi(T) \\
                                                        &= \sum_{\lambda_{X=x} \in T \in 2^U} \left( \prod_{Y \in \mathcal{V}} \mathrm{CPT}_Y \right)(T) \\
                                                        &= \sum_{(y_i)_{i=1}^n \in \prod_{i=1}^n \im Y_i} \Pr(x, y_1, \dots, y_n) \\
                                                        &= \Pr(X = x)
  \end{align*}
  by the following arguments:
  \begin{itemize}
  \item the proof of Theorem~1 by \citet{DBLP:conf/aaai/DudekPV20};
  \item if $\lambda_{X=x} \not\in T \in 2^U$, then $(\phi \cdot
    [\lambda_{X=x}])(T) = \phi(T) \cdot [\lambda_{X=x}](T \cap \{
    \lambda_{X=x} \}) = \phi(T) \cdot 0 = 0$;
  \item \cref{lemma:full_distribution};
  \item marginalisation of a probability distribution.
  \end{itemize}
\end{proof}
% TODO: need to make the proof nicer.

>>>>

%Many WMC algorithms have been proposed
%based primarily either on knowledge compilation
%\cite{DBLP:conf/ecai/Darwiche04,DBLP:conf/ijcai/OztokD15,DBLP:conf/ijcai/LagniezM17}
%or search \cite{DBLP:conf/aaai/SangBK05}.

\emph{Weighted model counting} (WMC), i.e., an extension of model counting
($\#\SAT$) that assigns a weight to every model \cite{DBLP:conf/aaai/SangBK05}
has emerged as the unifying inference mechanism across many (probabilistic)
domains. It has been applied to Bayesian networks
\cite{DBLP:conf/aaai/SangBK05,DBLP:books/daglib/0024906} and probabilistic
graphical models more generally \cite{DBLP:conf/ecsqaru/ChoiKD13}, probabilistic
logic programming languages \cite{DBLP:journals/tplp/FierensBRSGTJR15} and
probabilistic programs with discrete probability distributions
\cite{DBLP:journals/corr/abs-2005-09089}. Over the last fifteen years, WMC has
been extended and generalised in many ways, e.g., to handle continuous
probability distributions \cite{DBLP:conf/ijcai/BellePB15}, first-order
probabilistic theories
\cite{DBLP:conf/ijcai/BroeckTMDR11,DBLP:journals/cacm/GogateD16}, and infinite
domains \cite{DBLP:conf/aaai/Belle17}. Furthermore, the idea of weights in WMC
has been generalised from non-negative real numbers to an arbitrary semiring,
making it applicable to domains such as parsing, dynamic programming, constraint
programming, databases, and networks \cite{DBLP:journals/japll/KimmigBR17}.

The most well-studied version of WMC assigns weights to models based on
weights on literals, i.e., the weight of a model is the product of the weights
of all literals in it. This simplification is motivated by the fact that the
number of models scales exponentially with the number of atoms, so listing the
weight of every model is intractable. However, this also severely restricts what
probability distributions can be represented. A common way to overcome this
limitation is by adding more literals. While we show that this is always
possible, the rest of the paper demonstrates that it can be significantly more
efficient to encode weights in a more flexible way instead.

\Cref{sec:prelims} describes three equivalent perspectives on the subject based
on logic, set theory, and Boolean algebras. Furthermore, we describe the space
of functions on Boolean algebras and various operations on those functions. In
\cref{sec:wmc_as_measure}, WMC is introduced as the problem of computing the
value of a measure on a Boolean algebra. We show that not all measures can be
represented using literal-based WMC, but all Boolean algebras can be extended to
make any measure representable in such a manner.

In \cref{sec:bns}, we introduce \emph{conditional weights} on literals (in the
spirit of conditional probabilities) that have literal-based WMC as a special
case. As a particular example, we show how conditional weights can be used to
encode Bayesian networks. We prove the correctness of the encoding and show
how a state-of-the-art WMC solver ADDMC \cite{DBLP:conf/aaai/DudekPV20} can be
adapted to the new format. ADDMC is a recently-proposed algorithm for WMC based
on manipulating functions on Boolean algebras using an efficient representation
for such functions known as \emph{algebraic decision diagrams} (ADDs)
\cite{DBLP:journals/fmsd/BaharFGHMPS97}.

Finally, in \cref{sec:experiments}, we show how the new encoding results in
significantly improved inference speed on most instances, often by one or two
orders of magnitude. We explain the performance benefits by showing how the new
encoding has asymptotically fewer variables and ADDs.

%conditional weights allow us to represent
%probabilities in a manner that is both efficient and interpretable. Not having
%additional variables that are only needed to manipulate the representation into
%a specific format helps with both efficiency and interpretability. Similarly to
%Bayesian networks, the size of the representation scales with independence.

@@

\section{WMC as a Measure} \label{sec:wmc_as_measure}

@@

Finally, a measure $\mu\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ is
\emph{factorable} if there exists a factored weight function $\nu\colon 2^U \to
\mathbb{R}_{\ge 0}$ that induces $\mu$. In this formulation, WMC corresponds to
the process of calculating the value of $\mu_\nu(x)$ for some $x \in 2^{2^U}$
with a given definition of $\nu$.

@@

\section{Encoding Bayesian Networks Using Conditional Weights} \label{sec:bns}

@@

\correctness*
\begin{proof}
  Let $\mathcal{V} = \{ X, Y_1, \dots, Y_n \}$. Then
  \begin{align*}
    (\exists_U (\phi \cdot [\lambda_{X=x}]))(\emptyset) &= \sum_{T \in 2^U} (\phi \cdot [\lambda_{X=x}])(T) \\
                                                        &= \sum_{\lambda_{X=x} \in T \in 2^U} \phi(T) \\
                                                        &= \sum_{\lambda_{X=x} \in T \in 2^U} \left( \prod_{Y \in \mathcal{V}} \mathrm{CPT}_Y \right)(T) \\
                                                        &= \sum_{(y_i)_{i=1}^n \in \prod_{i=1}^n \im Y_i} \Pr(x, y_1, \dots, y_n) \\
                                                        &= \Pr(X = x)
  \end{align*}
  by:
  \begin{itemize}
  \item the proof of Theorem~1 by \citet{DBLP:conf/aaai/DudekPV20};
  \item if $\lambda_{X=x} \not\in T \in 2^U$, then $(\phi \cdot
    [\lambda_{X=x}])(T) = \phi(T) \cdot [\lambda_{X=x}](T \cap \{
    \lambda_{X=x} \}) = \phi(T) \cdot 0 = 0$;
  \item \cref{lemma:full_distribution};
  \item marginalisation of a probability distribution.
  \end{itemize}
\end{proof}
